---
title: "Harbor Seal Haul-out Behavior Analysis"
author: "Your Name"
format: html
editor: visual
---

## Introduction

This analysis explores the factors influencing harbor seal haul-out behavior, specifically examining how environmental variables and time of day affect the number of seals observed.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lme4)
library(lubridate)
```

## Data Processing

```{r data-processing}
# Read and process data
df <- read_csv("seals.csv") %>%
    rename(
        time = `Time of Day`,
        date = `Full Date`,
        seals = `#Seals`,
        temp1 = `Temperature(C)`,
        temp2 = `Temperature B (C)`,
        avgTemp = `Average Temp (C)`,
        tide = `Tide (ft)`
    )

# Create centered time variables
df <- df %>%
    mutate(
        datetime = mdy_hms(date),
        hourOfDay = hour(datetime) + minute(datetime) / 60,
        hourOfDay_centered = hourOfDay - mean(hourOfDay),
        hourOfDay2_centered = hourOfDay_centered^2
    )
```

## Exploratory Analysis

First, we visualize the daily pattern of seal haul-out behavior.

```{r daily-pattern}
df %>%
    group_by(time) %>%
    summarize(
        mean_seals = mean(seals, na.rm = TRUE),
        se_seals = sd(seals, na.rm = TRUE) / sqrt(n())
    ) %>%
    ggplot(aes(x = time, y = mean_seals)) +
    geom_col() +
    geom_errorbar(
        aes(
            ymin = mean_seals - se_seals,
            ymax = mean_seals + se_seals
        ),
        width = 0.2
    ) +
    theme_bw() +
    labs(
        x = "Time of Day",
        y = "Average Number of Harbor Seals Hauled Out",
        title = "Daily Pattern of Harbor Seal Haul-out Behavior"
    )
```

## Model Development

We explored several modeling approaches before settling on a log-transformed linear mixed model. Key considerations included:

1. High proportion of zeros in the data
2. Right-skewed distribution of non-zero counts
3. Temporal autocorrelation
4. Daily patterns in seal behavior

### Final Model

We fit a linear mixed model to log-transformed non-zero counts:

```{r final-model}
# Create dataset for analysis
df_counts <- df %>%
    filter(seals > 0) %>%
    mutate(
        log_seals = log(seals),
        date = as.Date(datetime)
    )

# Fit model
final_model <- lmer(
    log_seals ~ tide + avgTemp + hourOfDay2_centered +
        (1 | date),
    data = df_counts
)

# Model summary
summary(final_model)
```

### Model Diagnostics

```{r model-diagnostics}
# Basic diagnostic plots
par(mfrow = c(2, 2))
plot(final_model)

# Check for collinearity
car::vif(final_model)
```

## Results

The analysis revealed several key findings:

1. Time of day has a significant quadratic effect on seal numbers, with peak haul-out occurring around midday
2. Temperature shows a negative relationship with seal numbers
3. Tide level does not significantly affect seal numbers when they are present
4. There is substantial day-to-day variation in seal numbers (captured by the random effect)

### Effect Sizes

```{r effect-sizes}
# Extract and transform coefficients
coef_summary <- summary(final_model)$coefficients
coef_summary %>%
    as.data.frame() %>%
    rownames_to_column("Parameter") %>%
    knitr::kable(
        caption = "Model Coefficients",
        digits = 3
    )
```

## Discussion

[Add interpretation of results and biological significance]

## References

[Add relevant citations]

<iframe width="560" height="315" src="https://www.youtube.com/embed/8wkbgOxaHTM?si=iniq7cpYKSyEzBIU" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>

